Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-15 paper/intro.tex
finished through termination

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes,
where each read has a matching \emph{fulfilling} write,
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{wy2}{ry2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
% (In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
% range over shared memory references.)

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions which
are inconsistent with program order, such as:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \po{wx0}{wx1}
      \po[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}

@@

There are prior studies of relaxed memory that include sequential
composition and/or precise calculation of semantic dependencies.
\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.
They use step-indexing to account for
loops; we expect that the same approach could be applied here.
%
\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.
%
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.

Many of the most prominent models are based on speculative execution
\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
these models fail to validate compositional reasoning of temporal
properties---see their examples OOTA4 and OOTA5 (from
\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
can be understood in terms of the valid program transformations.  The
speculative models allow reads to be introduced, with subsequent case
analysis on the value read---effectively, this can turn one read into two,
with different conditional branches taken for the two copies of the read.
Our model invalidates this transformation.  In return, our model enjoys
compositionality for temporal safety properties.

>>>>

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes (black arrows),
where each read has a matching \emph{fulfilling} write (green arrows),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{wy2}{ry2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

% (In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
% range over shared memory references.)

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions which
are inconsistent with program order, such as:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \po{wx0}{wx1}
      \wk[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}

@@

There are few prior studies of relaxed memory that include sequential
composition and/or precise calculation of semantic dependencies.
\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.
They use step-indexing to account for
loops; we expect that the same approach could be applied here.
%
\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.
%
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational, whole-program models based
on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.

\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the speculative models
listed above all, including \cite{DBLP:conf/popl/KangHLVD17}, fail to
validate compositional reasoning of temporal properties---see their examples
OOTA4 and OOTA5 (from \citep{DBLP:journals/toplas/Lochbihler13}).  The
difference with our model can be understood in terms of the valid program
transformations.  The speculative models allow reads to be introduced, with
subsequent case analysis on the value read---effectively, this can turn one
read into two, with different conditional branches taken for the two copies
of the read.  Our model invalidates this transformation.  In return, our
model enjoys compositionality for temporal safety properties.
